---
title: "Tech Summary and Preliminary Estimates"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE)
library(tidyverse)
library(readr)
```

# Decomposing the headline inflation rate

I run a Fixed Effects OLS regression across our country panel to determine the contribution of core, food, and energy inflation rates to the headline harmonized CPI inflation (all data are from Eurostat). 

$CPI_{i,t} = \beta_{1}CORE_{i,t} + \beta_{2}FOOD_{i,t} + \beta_{3}ENERGY_{i,t}+  \delta_{i} + \gamma_{t} + \epsilon_{i,t}$

where $\delta_{i}$ are country fixed effects, $\gamma_{t}$ are annual fixed effects and $\beta_{j}$ reference the weight of each component in the headline inflation rate.

```{r}
infl <- read_rds('determinants of inflation/infl_decomp.rds')
```

Below are the results for the last four years of the sample (2019-2023):

```{r, fig.cap='Contribution of Core inflation, Food and Energy prices to headline inflation rates.', fig.width=15,fig.height=10}
infl %>% 
  ggplot(aes(x = date)) +
  geom_col(aes(y = contrib, fill = var), color = 'white') +
  geom_line(aes(y = CP00), linewidth = .75) + 
  facet_wrap(~country, scales = 'free'
  ) +
  theme_minimal() +
  theme(legend.position = 'bottom') +
  scale_fill_brewer(palette = 'Dark2', 'Contribution to\nheadline inflation', direction = -1) +
  labs(x = '',
       y = '')
```

Same plot with fixed "Y" axis:

```{r, fig.cap='Contribution of Core inflation, Food and Energy prices to headline inflation rates, fixed Y axis.',fig.width=15,fig.height=10}
infl %>% 
  ggplot(aes(x = date)) +
  geom_col(aes(y = contrib, fill = var), color = 'white') +
  geom_line(aes(y = CP00), linewidth = .75) + 
  facet_wrap(~country, scales = 'free_x'
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = 'Dark2', 'Contribution to\nheadline inflation', direction = -1) +
  labs(x = '',
       y = '')
```

Potential follow-up:

Mediation analysis (/ an instrumental variable approach) could show if there is any inter-play among the rates. It is not too unreasonable to believe that government measures, such as price caps were instead realized in growth of non-targeted goods. For example, price caps on food or fuel items - resulting in revenue losses - were recouped by firms via increasing the prices of items found in the core CPI basket. This could be done by regressing each component on the other components and accounting for the indirect effects.


# Real rates, equilibrium and policy stance

Here I first calculate real interest rates as $r_{t} = i_{t} - \pi_{t+1}$, where $r_{t}$ is the real rate, $i_{t}$ is the nominal rate and $\pi_{t}$ is the annualized CPI inflation rate. The graph below shows the real rates across the countries in the sample:

```{r}
real_rate <- read_rds('real rate/real_rate.rds')
```

```{r, fig.cap= 'Real rates across countries.',fig.width=15,fig.height=10}
real_rate %>% 
  ggplot(aes(x = date, y = r)) +
  geom_line(linewidth = .75)+
  facet_wrap(~country, scales = 'free_x') +
  theme_minimal() +
  labs(x = '', 
       y = '')
```

To separate the natural rate from cyclical fluctuations, I use two variations of the HP filter, a One-sided, and a Two-sided version. The two approaches yield similar trends, with the one sided (naturally) showing more fluctuations. The trend component estimated with the filters can be considered as the equilibrium real interes rate, these can be seen in the graph below:
```{r, fig.cap='Natural real interest rates.',fig.width=15,fig.height=10}
real_rate %>% 
  select(country, date, r_trend1, r_trend2) %>% 
  gather(filter, value, starts_with('r')) %>% 
  mutate(filter = case_when(filter == 'r_trend1' ~ 'One-sided HP',
                            filter == 'r_trend2' ~ 'Two-sided HP')) %>% 
  ggplot(aes(x = date, y = value, color = filter)) +
  geom_line(linewidth = .75) +
  facet_wrap(~country, scales = 'free') +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = 'bottom') +
  labs(x = '',
       y = '') +
  scale_color_manual(breaks = c('One-sided HP', 'Two-sided HP'),
                     values = c('red', 'darkblue'))

```
The graphs here show a historically declining equilibrium rate up to 2019/2020, where we can observe a reversion in this trend.

The deviations from the trend, i.e. the cyclical components thus can be used to asses policy stance:
```{r 'Policy stance - deviation from the equilibrium.',fig.width=15,fig.height=10}
real_rate %>% 
  select(country, date, r_cycle1, r_cycle2) %>% 
  gather(filter, value, starts_with('r')) %>% 
  mutate(filter = case_when(filter == 'r_cycle1' ~ 'One-sided HP',
                            filter == 'r_cycle2' ~ 'Two-sided HP')) %>% 
  ggplot(aes(x = date, y = value, color = filter)) +
  geom_line(linewidth = .75) +
  facet_wrap(~country, scales = 'free') +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = 'bottom') +
  labs(x = '',
       y = '') +
  scale_color_manual(breaks = c('One-sided HP', 'Two-sided HP'),
                     values = c('red', 'darkblue'))

```
All graphs here agree that policy stance was following a very loose monetary policy around 2-8 percent below the equilibrium, at the time the war shock hit the economies. The graphs also show the fast response of the regional central banks, as the real rate gaps rose sharply in the following months.
