---
title: "Macroeconomic aggregates used in our analyisis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval = TRUE)
library(tidyverse)
```

This markdown file showcases the data series used in our book chapter, indicating our data sources.

# Inflation rates

Monthly CPI inflation rates retrieved, from the IMF IFS database.

```{r, fig.cap = 'Inflation rates in the CEE countries'}

read_rds('input/inflation.rds') %>% 
  group_by(country) %>% 
  mutate(cpi_ann = cpi - lag(cpi, 12)) %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = cpi_ann)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# GDP

Quarterly Real GDP in national currency, seasonally adjusted, retrieved from the IMF IFS database.

```{r, fig.cap = 'GDP in the CEE countries'}

read_rds('input/gdp.rds') %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = gdp)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# Industrial Production

Monthly industrial production index, seasonally adjusted using X13-ARiMA, retrieved from the IMF IFS database.

```{r, fig.cap = 'Industrial production indices in the CEE countries'}

read_rds('input/indpro.rds') %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = indpro)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```


# Interest Rates

Monthly short run interest rates, measured as policy rates retrieved from the IMF IFS database, or 3-month interbank rates retrieved from the OECD Main Main Financial and Economic Indicators database.

```{r, fig.cap = 'Short interest rates in the CEE countries'}

read_rds('input/intrate.rds') %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = r)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# Core Inflation Rates

Monthly core inflation rates measured as CPI inflation less energy and food inflation, retrieved from Eurostat.

```{r, fig.cap = 'Core inflation rates in the CEE countries'}

read_rds('input/core_infl.rds') %>%
  arrange(country, date) %>% 
  group_by(country) %>% 
  mutate(core_infl = core_infl - lag(core_infl, 12)) %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = core_infl)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# Exchange Rates

Monthly exchange rates, measured as Real Broad Effective Exchange Rates, retrieved from the BIS EER database.

```{r, fig.cap = 'Exchange rates in the CEE countries'}

read_rds('input/reer.rds') %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = reer)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# Wage rates

Quarterly real wage rate indices, seasonally and calendar adjusted, measured as wage growth of sectors B-S, adjusted for inflation using GDP deflator, retrieved from Eurostat. The series are interpolated from quarterly to monthly frequency using spline interpolation.

```{r, fig.cap = 'Wage rates in the CEE countries'}

read_rds('input/wage.rds') %>% 
  arrange(country, date) %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = wage)) +
  geom_line() +
  facet_wrap(~country, scales = 'free') +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# Energy Prices

Monthly global energy price index, retrieved from the IMF PCPS database.

```{r, fig.cap = 'Global Energy Price Index'}

read_rds('input/enprice.rds') %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = enprice)) +
  geom_line() +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```

# Geopolitical Risk Index

Monthly measurement of geopolitical tensions, created by Dario Caldara and Matteo Iacoviello using automated text searches, retrieved from policyuncertainty.com

```{r, fig.cap = 'Global Geopolitical Risk Index'}

read_rds('input/gpr.rds') %>% 
  filter(year(date) >= 2000) %>% 
  ggplot(aes(x = date, y = gpr)) +
  geom_line() +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1)) +
  labs(x = '',
       y = '')

```









